Galoissche Kennzeichnung von Funktionenkörper \\ vom Transzendenzgrad $1$ über Zahlkörper

### 2017-06-03 Master Seminar/article.tex
More reduction

We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
where $S$ ranges over all finite subsets of $K$. Let $v,w\in\operatorname{Spv}(K)$. We have:
\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
Thus $v$ is a closed point if and only if it has no non-trivial specializations.
\end{definition}

\begin{lemma}\label{rank-dimension-inequality}
Let $v\in\operatorname{Spv}(K)$. Then we have:
\begin{enumerate}[(i)]
\item $\operatorname{rank}(v)\leq \operatorname{rrank}(v)$
\item $\operatorname{rrank}(v)+\dim k(v)\leq\dim(K)$
\end{enumerate}
Moreover, if equality holds in (ii) and $K$ is finitely generated, then $k(v)$ is finitely generated as well, and $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module.
\end{lemma}

\begin{definition}
We define:
\begin{align*}
\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \} \\
\operatorname{Spv}_\text{rd}(K) &= \{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rank}(v)=\operatorname{rrank}(v) \}
\end{align*}
Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}(K)$ are called \textit{rank-defectless}\index{valuation!rank-defectless}. For $\bullet\in\{\text{d},\text{rd}\}$ we set:
\[\operatorname{Spv}_\bullet^{(i)}(K)=\{v\in\operatorname{Spv}_\bullet(K)\mid \operatorname{rrank}(v)=i \} \]
\end{definition}

\section{Main Lemma}

@@

\begin{lemma}\label{2.5}
Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$. Set $k=k(w)$ and $\Gamma = \Gamma_w$. Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$. Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$. Then:
\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [k:\mathbb{F}_p] \text{ and } \dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2\\
0, & \text{otherwise}
\end{cases} \]
If $L/K$ is an algebraic extension and $\mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \mathbb{Z}/\ell\mathbb{Z}$ holds, we have:
\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K]\\
0, & \text{otherwise}
\end{cases} \]
\end{lemma}

\begin{proposition}\label{2.6}
Let $K_0/K$ be an algebraic extension.
\begin{enumerate}[(i)]
\item Suppose there exists a prime number $\ell$ with $\ell^\infty\nmid[K_0:K]$ and such that for every algebraic extension $L/K_0$ we have: 
\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K_0]\\
0, & \text{otherwise}
\end{cases} \]
Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is a defectless valuation.
\end{enumerate}
\end{proposition}

\begin{proof}
\begin{enumerate}[(i)]
\item Since $\mathrm{H}^3(K_0, \mathbb{Z}/\ell\mathbb{Z})\neq 0$ by assumption, according to Lemma \ref{2.4} we can find a valuation $w_0\in|\operatorname{Spv}(K_0)|$ such that:
\[ \mathrm{H}^3((K_0)_{w_0}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0 \]
By requirement $\ell^\infty \nmid [(K_0)_{w_0} : K_0]$. By replacing $K_0$ by $(K_0)_{w_0}$, we can assume that $w_0$ is henselian.

Assume that $w = w_0|_K$ is not henselian. Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique extension to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.

Let $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$. Then $(K_0)_{w_1}$ is the composite of $K_0$ and $K_w$ and we have $\ell^\infty\nmid [(K_0)_{w_1} : K_w]$.

Since $K_0$ is henselian with respect to $w_0$, a conjugate $\tau(K_w)$ for some $\tau\in G_K$ of $K_w$ lies in $K_0$, hence $\ell^\infty\nmid [K_w : K]$ as well. Consequently, $\ell^\infty \nmid [(K_0)_{w_1} : K]$ and therefore $\ell^\infty \nmid [(K_0)_{w_1} : K_0]$. By requirement, $\mathrm{H}^3((K_0)_{w_1}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0$.

Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$. A contradiction to $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$.
\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds. According to Lemma \ref{2.5} we have:
\[\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)\qedhere\]
\end{enumerate}
\end{proof}

@@

\begin{proof}
Let $U = X -  S$ and $P_1\in U$ be a closed point. We set:
\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
Therefore, $C_1$ is a free abelian group of finite rank. If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done. Otherwise, let $x\in C_1,\ x\neq 1$. Then there exists an $a\in k^\times$ with the following properties:
\begin{enumerate}[(i)]
\item There exists a point $P_2\in X_0 - (S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
\item $a$ is not a root of unity.
\end{enumerate}
Clearly, (i) and (ii) hold for almost all $a\in k^\times$. We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$. $C_2$ is free abelian of finite rank as well. By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$. By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective.
\end{proof}

\begin{definition}
For each admissible $S\subset X_0$ and $T_1\subset X_0 -  S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$. Therefore,  we can find a finite subset $T\subset X_0 -  S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective. Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
\end{definition}

@@

\begin{proof}
By Lemma \ref{5.4}, the following maps are injective for all admissible subsets $S\subset X_0$:
\[ r_S:K^S \to\prod_{P\in X_0 -  S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
Now it's evident that $\varphi$ is unique. For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
\[ (\star) \qquad \begin{tikzcd}
K'^{S'} \rar["\varphi^{S,T}", dashed] \dar["r_{S',T'}"'] & K^S \dar["r_{S,T}"]\\
\displaystyle \prod_{P\in T}k(\sigma P)^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
\end{tikzcd} \phantom{\qquad (\star)} \]
The vertical maps are injective by definition. Thus, it suffices to show that the following composition is trivial:
\[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(\sigma P)^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
Consider the commutative diagram which results from tensoring $(\star)$ with $\mathbb{Z}/n\mathbb{Z}$:
\[ \begin{tikzcd}
K'^{S'}/n \rar[dashed]\dar & K^S/n\dar\\
\displaystyle \prod_{P\in T}k(\sigma P)^\times/n \rar & \displaystyle \prod_{P\in T}k(P)^\times/n
\end{tikzcd} \]
According to Lemma~\ref{4.3} and Lemma~\ref{4.5} (ii) this diagram can now be completed by $\varphi_n: K'^{S'}/n\to K^S/n$. This implies $\varrho/n=0$, i.e. $\operatorname{im}(\varrho)\subset n\operatorname{coker}(r_{S,T})$ for every $n\in\mathbb{N}$. It follows $\varrho = 0$. By symmetry, $\varphi^{S,T}$ is an isomorphism.

$\varphi^{S,T}$ actually independent of $T$, thus we can omit the $T$ from its notation. Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P \Big)\circ r_{S'} \]
In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$.

We have $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$. Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$. 
\end{proof}

@@

\begin{remark}
For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$. If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$ according to Remark \ref{3.3} and Step 1. By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
\end{remark}

>>>>

\[|\operatorname{Spv}(K)|=\{v\in\operatorname{Spv}(K)\mid \dim k(v)=0\}\]
\end{definition}

\begin{lemma}\label{rank-dimension-inequality}
Let $v\in\operatorname{Spv}(K)$. Then we have $\operatorname{rrank}(v)+\dim k(v)\leq\dim(K)$. If equality holds and $K$ is finitely generated, then $k(v)$ is finitely generated as well, and $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module.
\end{lemma}

\begin{definition}
We define:
\begin{align*}
\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \}
\end{align*}
Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}.
\[\operatorname{Spv}_\text{d}^{(i)}(K)=\{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rrank}(v)=i \} \]
\end{definition}

@@

\begin{lemma}\label{2.5}
Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$. Then $k(w)$ is an algebraic extension of a finite field $\mathbb{F}_p$. Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$. Then:
\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [k(w):\mathbb{F}_p] \text{ and } \dim_{\mathbb{F}_\ell}(\Gamma_w / \ell\Gamma_w) = 2\\
0, & \text{otherwise}
\end{cases} \]
If $L/K$ is an algebraic extension and $\mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \mathbb{Z}/\ell\mathbb{Z}$ holds, we have:
\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K]\\
0, & \text{otherwise}
\end{cases} \]
\end{lemma}

\begin{proposition}\label{2.6}
Let $K_0/K$ be an algebraic extension. Suppose there exists a prime number $\ell\neq 2$ with $\ell^\infty\nmid[K_0:K]$ and such that for every algebraic extension $L/K_0$ we have: 
\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K_0]\\
0, & \text{otherwise}
\end{cases} \]
Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
\end{proposition}

@@

@@

\begin{proof}
Let $S\subset X_0$ be admissible and $T\subset X_0-S$ such that $r_{S,T}$ and $r_{S',T'}$ are injective. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
\[ (\star) \qquad \begin{tikzcd}
K'^{S'} \rar["\varphi^{S,T}", dashed] \dar["r_{S',T'}"'] & K^S \dar["r_{S,T}"]\\
\displaystyle \prod_{P\in T}k(\sigma P)^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
\end{tikzcd} \phantom{\qquad (\star)} \]
The vertical maps are injective by definition. Thus, it suffices to show that the following composition is trivial:
\[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(\sigma P)^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
Consider the commutative diagram which results from tensoring $(\star)$ with $\mathbb{Z}/n\mathbb{Z}$:
\[ \begin{tikzcd}
K'^{S'}/n \rar[dashed]\dar & K^S/n\dar\\
\displaystyle \prod_{P\in T}k(\sigma P)^\times/n \rar & \displaystyle \prod_{P\in T}k(P)^\times/n
\end{tikzcd} \]
According to Lemma~\ref{4.3} and Lemma~\ref{4.5} (ii) this diagram can now be completed by $\varphi_n: K'^{S'}/n\to K^S/n$. This implies $\varrho/n=0$, i.e. $\operatorname{im}(\varrho)\subset n\operatorname{coker}(r_{S,T})$ for every $n\in\mathbb{N}$. It follows $\varrho = 0$. By symmetry, $\varphi^{S,T}$ is an isomorphism.

$\varphi^{S,T}$ actually independent of $T$, thus we can omit the $T$ from its notation. Now if $T$ ranges over all $S$-admissible subsets of $X_0$, we can replace $T$ by $X_0 -  S$ in $(\star)$. We can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$. 
\end{proof}

@@

\begin{remark}
For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$. By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
\end{remark}
